Application of Micro-Rotors to Induce Crystal Nucleation: Challenges, Takeaways, and Future Prospects\\ PhD Thesis

### 2025-07-11 Latex_files/Thesis/chapters/chapter 5/main.tex
updates to QPD chapter

\section{Introduction}

As demonstrated in Chapter 3, the motion and behaviour of a dimer 
is significantly different compared to that of a traditional sphere.
While this is simple enough to demonstrate in a simulation, where the 
position and orientation can be computed using the finite differences
method, from an experimental perspective these details can go unnoticed.
In order to accurately capture the dynamics of a trapped sphere, a 
position detection system is typically used. This can capture the 
translational motion of a trapped particle but relating the signals
produced to the rotational motion is not as straight forward.

@@

\subsection{Position Detection methods}
There are 3 common methods of position detection: video-analysis, lateral-effect position sensing, and photodiodes. The former is 
relies purely on cameras to track particle displacements; whereas 
the latter two are examples of focal-plane interferometry. 
Interferometry utilises the interference between the scattered and un-scattered light to measure displacement. Here we 
provide a brief outline of their applications, and limitations.

Video analysis is ideally suited for multiple traps or situations where 
precision is not the top priority. In order for video analysis to 
match the accuracy of back-focal plane interferometry methods requires 
the camera's frame rate to exceed $1\ kHz$ which can be difficult to 
achieve while maintaining a decent resolution \cite{Gibson2008}. In 
comparison off the shelf back-focal plane detectors can achieve temporal 
resolutions anywhere from $10-100\ kHz$ \cite{BergSoerensen2004}.

A quadrant photo diode (QPD) is a frequently used position detection 
system for optical tweezers due to their high sampling rate, high 
degree of precision, and ease of set up. The QPD is constructed of 
four photo diodes assembled in a quadrant formation, when a particle 
is trapped the interference pattern produced is focused onto the QPD, 
with the maximum intensity mapping to the particle's centre of mass. 
By summing the voltages of the horizontal and vertical quadrants together 
the particle's centre of mass is tracked in the $x-y$ plane. Axial 
displacement can be estimated by observing the change in the total 
voltage of the QPD. The outputted signal gives an indication of the 
particle's relative displacement from the beam focus, but in order to 
convert the signal to distance units the trap needs to be calibrated 
(assuming a linear response curve).

\begin{figure}[h!]
	\centering
	\includegraphics[height=\linewidth, angle=270]{QPD_Lateral_effect.pdf}
	\caption{Comparison between QPD and Lateral effect photodiodes.
		The four quadrants of a QPD (left) experience different photocurrents
		based on the total intensity of light incident on each section 
		(labelled A, B, C, D). 
		Whereas a Lateral effect sensor (right) uses the resistive properties
		of the photodiode surface to vary the create different photocurrents 
		passing through the anodes A, B, C, and D.}
\end{figure}

A lateral-effect sensor works using a single cell analogous to the focal plane of the trapping beam. The four corners of the sensor act as anodes connected to a base plate cathode, as the beam moves across the surface of the detector each anode will experience a different photocurrent depending 
on how close the centre of the interference pattern is to each anode. 
The advantage of a lateral effect detector is that the linear regime 
is much larger than a QPD making it much better for monitoring the 
position of a trapped particle. However, Lateral-effect sensors are 
often limited in their spacial resolution due to high signal-to-noise 
ratios, requiring a high intensity of light on the sensor in order to 
get a clean signal. As a result, most optical force measurements are 
conducted using a QPD as opposed to a lateral-effect sensor, as often 
the displacement is small enough that the signal-displacement curve 
can be considered linear.

The above informs how one can - to their best ability - measure small 
rotational movements in a trajectory. Relying on video analysis is 
unreliable as the frame rate required to accurately measure the 
orientation results in a loss of resolution. Likewise, simply measuring
the total intensity provides no means of separating out rotational and 
translational motion. Similar to Chapter 4, we need to utilise contrasting signals to estimate the change in orientation, and see if at all it is possible to distinguish between translational and rotational motion.

\section{Modelling the QPD response signal of a non-spherical particle}
In order to simulate a typical experimental set up with a QPD 
installed as a position detection system we need to evaluate 
the total magnitude of the electric field incident on the 
photo-diode surface. While trapping a micro-particle, the 
scattered and incident fields combine together and interfere 
with one another. These fields are collected by a condenser 
lens in the far field limit and are focused onto the QPD 
surface, the total intensity can be evaluated as:
\begin{align}
	I(x,y) = \epsilon_0c\left|
	\begin{bmatrix} 
		E_{i,x}(x,y)+E_{s,x}(x,y) \\ 
		E_{i,y}(x,y)+E_{s,y}(x,y) \\ 
		E_{i,z}(x,y)+E_{s,z}(x,y)
	\end{bmatrix} \right|^2 \times step(\theta_{NA}-\theta(x,y))
\end{align}

@@

While translational motion has no knock on effects to the QPD signal, 
rotational changes are double counted by both the incident and total 
field. This results in rotational motion being biased in the QPD signal 
- even when collecting signals from isotropic scatterers. To prevent 
this we need to rotate the total field via the inverse rotation matrix
of the dimer.

To confirm that our method is producing accurate results, we ran a 
comparison between our simulative QPD and the results from \cite{Rohrbach2002}. Where a $300\ nm$ diameter sphere is scanned 
across the path of a focused Gaussian beam ($\lambda=1064\ nm$, $NA=1.2$), the sphere has a refractive index of 1.57 and is suspended in water ($n_{med}=1.33$) and the condenser lens has its numerical aperture
set to 0.5 ($\theta_{max} = 30^\circ$). Scanning across all three primary axis produced the following response curve:
\begin{figure}[h!]
	\begin{subfigure}{0.475 \linewidth}
		\subcaption{}
		\includegraphics[width=\linewidth]{QPD_axial_tests.png}
	\end{subfigure}
	\begin{subfigure}{0.475 \linewidth}
		\subcaption{}
		\includegraphics[width=\linewidth]{QPD_lat_tests.png}
	\end{subfigure}
	\caption{Comparison between QPD response signal versus work conducted by 
		Rohrbach, single sphere ($r = 150\ nm$, $n=1.57$) is scanned by a $1064\ nm$ laser and the QPD signal recorded. Solid lines represent the signal produced by QPD using \textit{ott} and points represent the signal response collected from \cite{Rohrbach2002}.}
	\label{fig:Rohrbach}
\end{figure}

@@

\section{Characterisation of asymmetric dimer dynamics via PSD analysis}
As discussed in Sec.~\ref{sec:simulated_QPD}, one of the methods 
developed to work in conjunction with \cite{Vigilante2020} 
is a simulated quadrant photo diode for as a position 
detection system. While it is possible to extract all of 
the relevant dynamics from a simulation, from an experimental
standpoint characterising those dynamics is dependent on what
experimental techniques are used. Translational motion can be
characterised via a position detection system but angular 
motion is far more difficult to detect, let along characterise.

>>>>

\section{Introduction}
As demonstrated in Chapter 3, the motion and behaviour of a dimer 
is significantly different compared to that of a traditional sphere.
While this is simple enough to demonstrate in a simulation, where the 
position and orientation can be computed using the finite differences
method, from an experimental perspective these details can go unnoticed.
In order to accurately capture the dynamics of a trapped sphere, a 
position detection system is typically used - the most common of which is 
a Quadrant Photo Diode (QPD). A QPD sends a signal measuring 
the voltage difference between 4 photo diode quadrants. For purely 
translational motion the diodes can be split into neighbouring pairs 
(i.e. left and right, top and bottom - see Fig~\ref{fig:QPD_diagram})
to estimate the total displacement made by the target particle. 
However the same method cannot be applied to stochastic rotational 
motion.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Position Detection methods}
Here we discuss how one can record a trajectory using a QPD and 
use it to interpret the particle's real time motion. The QPD is 
constructed of four photo diodes assembled in a quadrant formation, 
when a particle is trapped the interference pattern produced is 
focused onto the QPD, with the maximum intensity mapping to the 
particle's centre of mass. By summing the voltages of the horizontal 
and vertical quadrants together the particle's centre of mass is 
tracked in the $x-y$ plane. Axial displacement can be estimated 
by observing the change in the total voltage of the QPD. The 
outputted signal gives an indication of the particle's relative 
displacement from the beam focus, but in order to convert the 
signal to distance units the trap needs to be calibrated (assuming 
a linear response curve).

\begin{figure}[h!]
	\centering
	\includegraphics[height=0.5\linewidth, angle=270]{QPD_diagram.png}
	\caption{Diagram depicting a typical QPD interface. The four 
		quadrants of a QPD experience different photocurrents based 
		on the total intensity of light incident on each section 
		(labelled A, B, C, D).}
	\label{fig:QPD_diagram}
\end{figure}

Now while the signal returned by the QPD is proportional to the particle's
displacement it cannot be directly interpreted as such until converted
from voltage units to displacement units. This can be done by taking the 
ratio of the expected diffusion coefficient $D_{exp}$ and the diffusion 
coefficient calculated by calibration $D_{cal}$.
\begin{align}
	\label{eq:conversion_factor}
	D_{exp} = \frac{k_BT}{6\pi\eta a} \Rightarrow Conversion\ Factor \ [m/V]= \sqrt{\frac{D_{ext}}{D_{cal}}}
\end{align}

The former depends on the exact size and shape of the target particle, 
using image analysis one can get a fairly accurate estimation of the particles radius (a). The latter can be determined from fitting the QPD signal to a given model, for computing the diffusion coefficient 
computing the auto correlation function, the mean square displacement, 
or taking the power spectrum are used. The former two are ideal for 
simple position detection systems, where the QPD reports the signal at 
each time step, however they often require longer recording times in 
order to reach an ideal accuracy. More sophisticated QPD systems can calculate the power spectra of an incoming signal and immediately fit 
it to a power spectrum. Once the conversion factor is known any further 
QPD signals can be converted immediately into displacement units to monitor
the position of the target particle, the spacial resolution being  
limited by the accuracy of the calibration method. A full tutorial in 
maximising the accuracy of the power spectral method is provided by 
Berg and Sorensen \cite{BergSoerensen2004}

For our consideration we will be relying on the power spectrum method 
to characterise the motion of the trapped particle. Firstly because
it is sometimes integrated into QPD systems. And secondly, because 
on of the main advantages of the power spectrum method is its shorter 
calibration times, and the ability to filter out large sources of signal 
noise.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Modelling the QPD response signal of a non-spherical particle}
In order to simulate a typical experimental set up with a QPD 
installed as a position detection system we need to evaluate 
the total magnitude of the electric field incident on the 
photo-diode surface. While trapping a micro-particle, the 
scattered and incident fields combine together and interfere 
with one another. These fields are collected by a condenser 
lens in the far field limit and are focused onto the QPD 
surface, the total intensity can be evaluated as:
\begin{align}
	I(x,y) = \epsilon_0c\left|
	\begin{bmatrix} 
		E_{i,x}(x,y)+E_{s,x}(x,y) \\ 
		E_{i,y}(x,y)+E_{s,y}(x,y) \\ 
		E_{i,z}(x,y)+E_{s,z}(x,y)
	\end{bmatrix} \right|^2 \times step(\theta_{NA}-\theta(x,y))
\end{align}

@@

While translational motion has no knock on effects to the QPD signal, 
rotational changes are double counted by both the incident and total 
field. This results in rotational motion being biased in the QPD signal 
- even when collecting signals from isotropic scatterers. To prevent 
this we rotate the total field via the inverse rotation matrix of the 
dimer.

To confirm that our method is producing accurate results, we ran a
comparison test between our simulative QPD and the results from \cite{Rohrbach2002}. Where a $300\ nm$ diameter sphere is scanned 
across the path of a focused Gaussian beam ($\lambda=1064\ nm$, $NA=1.2$), the sphere has a refractive index of 1.57 and is suspended in water ($n_{med}=1.33$) and the condenser lens has its numerical aperture
set to 0.5 ($\theta_{max} = 30^\circ$). Scanning across all three primary axis produced the following response curve:
\begin{figure}[h!]
	\begin{subfigure}{0.475 \linewidth}
		\subcaption{}
		\includegraphics[width=\linewidth]{QPD_axial_tests.png}
	\end{subfigure}
	\begin{subfigure}{0.475 \linewidth}
		\subcaption{}
		\includegraphics[width=\linewidth]{QPD_lat_tests.png}
	\end{subfigure}
	\caption{Comparison between QPD response signal versus work conducted by 
		Rohrbach, single sphere ($r = 150\ nm$, $n=1.57$) is scanned by a $1064\ nm$ laser and the QPD signal recorded. Solid lines represent the signal produced by QPD using \textit{ott} and points represent the signal response collected from \cite{Rohrbach2002}.}
	\label{fig:Rohrbach}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Characterisation of asymmetric dimer dynamics via PSD analysis}
While it is possible to extract all of the relevant dynamics from a simulation, from an experimental standpoint characterising those 
dynamics is dependent on what experimental techniques are used. 
Translational motion can be characterised via a position detection 
system but angular motion is far more difficult to detect, let along characterise.
